Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-06-28 src/tex/ms.tex
Update

The LIGO-VIRGO-KAGRA (LVK) collaboration
~\citep{LIGOScientific:2014pky,LIGOScientific:2021usb,LIGOScientific:2021djp,
VIRGO:2014yos,KAGRA:2020tym} recently started their fourth observational
run on May 26, 2023. Impressively, they are
expected to double the total number of observed binary black holes (BBHs)~\citep{abbott2020prospects}.
Morover, the improved sensitivity also implies that we expect to
detect individual events with a higher signal-to-noise ratio (SNR) than ever before. This means
we can resolve more features in the signal, therefore putting more
stringent requirements on the accuracy of our waveform model \citep{purrer2020gravitational, hu2022assessing}.

@@

The SXS NR waveforms are given as time-series strain.
Since IMRPhenomD is modeled in the frequency domain, we need to Fourier transform the simulation results
in order to compute the mismatch, \eqref{eq:mismatch}.
For this, we taper the time-series using Tukey window~\citep{usman2016pycbc}\footnote{
	Specifically, we choose $\alpha=2t_{\mathrm{RD}}/T$, where
	$t_{\mathrm{RD}}$ is the duration of ringdown (maximum amplitude to the end of the strain) and $T$ is the duration of the
	entire GW strain. 
%	\te{What is $t_{\mathrm{RD}}$? and how is it computed?}
} before using standard FFT routines to compute the Fourier transform.

@@

To evaluate how well the optimization procedure generalizes to waveforms that are
not in the training set, we evaluate the mismatch between the fine-tuned model
and an additional 536 NR waveforms in the SXS catalog i.e., the validation set.
We select waveforms that share the same part of the parameter space with the training
set, i.e., waveforms with negligible eccentricity (${e<2\times10^{-3}}$)
and precession (${\chi_{x,y}<5\times10^{-3}}$). Figure~\ref{fig:intrin_space}
shows how the training and validation waveforms are distributed in the
$q-\chi_{\rm PN}$ space.

@@

\begin{figure}[t]
	\script{q148.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/q148.pdf}
	\caption{Distributions of mismatches before and after optimization using the 
	$\mathcal{L}_{\mathrm{ave}}$ loss function. Mismatches are calculated using the training waveforms listed in 
	Tab.~\ref{tab:q148} and	are weighted with a constant PSD. 
	The dashed lines represent the median of the distributions, which decreased by 45.3\% during optimization.}
	\label{fig:q148}
\end{figure}
\begin{figure}[t]
	\script{q148_q1248_compare.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/q148_q1248_compare.pdf}
	\caption{Distributions of $\log_{10}$ difference in mismatch. The red
	distribution uses the training waveforms listed in
	Tab.~\ref{tab:q148} while the green distribution uses waveforms
	listed in Tab.~\ref{tab:q148}~and~\ref{tab:q1248}. Mismatches are calculated
	using a constant PSD with the loss function $\mathcal{L}_{\mathrm{ave}}$. 
	The dashed lines represent the median of the distributions, which decreased \te{by an additional 10.4\% during optimization with the additional data.}}
	\label{fig:q148_q1248_compare}
\end{figure}

Note that the performance of the IMRPhenomD model was initially tested using the
{\zdethp} weighted mismatch.\footnote{
	Note, however, that the mismatch was never directly used during the calibration process~\citep{khan2016frequency, khan2019phenomenological}.
} We would therefore like to examine whether using the {\zdethp} PSD in our loss function
could lead to an improved to mismatch. Performing the optimization again, we find no significant difference
between the results using the two PSDs in the distribution of mismatches.

@@

We divided the parameter space into four regions to analyze the effect of the
recalibration procedure on each region separately
(Fig.~\ref{fig:all_quadrants}~and~\ref{fig:ps_q148_quadrant}). The training waveforms used for fitting in
this scenario are listed in Tab.~\ref{tab:quadrants} and the loss functions 
are calculated using a simple average of the mismatches, $\mathcal{L}_{\mathrm{ave}}$. The top-left and
bottom-right regions have limited data for $q>4$, hence the result is only valid up to $q\leq4$ and we take waveforms with $q\leq4$ as validation waveforms in these two regions. From top panel of Fig.~\ref{fig:ps_q148_quadrant}, we observe that
equal-spin waveforms lying on the diagonal improve the most. 
\kl{Both top-right and bottom-left region have good performance comparing with the original model. Especially for the bottom-right region, where the improvement is significantly better than optimizing with all waveforms. This suggests the ansatz are well-fitted to this part of the parameter space.
Waveforms above the diagonal also give good improvements when comparing with the original model and it has a similar improvement as optimizing all waveforms at once. Notice optimizing over all waveforms gives a slightly better improvement. This shows that the ansatz can achieve a better fit using training waveforms with similar spins, even for waveforms with non-equal spin parameters.}
%The improvement above the diagonal is also significantly more than when optimizing using
%all waveforms at once. This can be seen clearly in the top panel of Fig.~\ref{fig:all_quadrants}, where we see
%that the optimized histogram shifts to lower mismatches.
\footnote{
	Note that the bottom-left panel of Fig.~\ref{fig:all_quadrants} specifically shows the
	distribution of mismatches for the top-left quadrant of Fig.~\ref{fig:ps_q148_quadrant}.
} 
On the other hand, the improvement below the diagonal is negligible, indicating that our method struggles to
further improve over the original IMRPhenomD waveform. 
Interestingly, we can see in the bottom-right region that the waveform does not show a significant change
to the overall mismatch (see bottom-right panel of Fig.~\ref{fig:all_quadrants}). 
This suggests that the IMRPhenomD waveform ansatz fits waveforms with
$\chi_1>0$ and $\chi_2<0$ less well. 
\kl{Overall, the improvement for both optimization schemes are similar (bottom panel of Fig.~\ref{fig:ps_q148_quadrant}). Although split region optimization uses more training waveforms, some of the waveforms, e.g. opposite spins, hinders the optimization procedure and hence gives a slightly worse result when comparing with optimization over all waveforms.}

>>>>

The LIGO-VIRGO-KAGRA (LVK) collaboration
~\citep{LIGOScientific:2014pky,LIGOScientific:2021usb,LIGOScientific:2021djp,
VIRGO:2014yos,KAGRA:2020tym} recently started their fourth observational
run on May 26, 2023. Impressively, they are
expected to double the total number of observed binary black holes (BBHs)~\citep{abbott2020prospects}.
Moreover, the improved sensitivity also implies that we expect to
detect individual events with a higher signal-to-noise ratio (SNR) than ever before. This means
we can resolve more features in the signal, therefore putting more
stringent requirements on the accuracy of our waveform model \citep{purrer2020gravitational, hu2022assessing}.

@@

The SXS NR waveforms are given as time-series strain.
Since IMRPhenomD is modeled in the frequency domain, we need to Fourier transform the simulation results
in order to compute the mismatch, \eqref{eq:mismatch}.
For this, we taper the time-series using Tukey window~\citep{usman2016pycbc}\footnote{
	Specifically, we choose $\alpha=2t_{\mathrm{RD}}/T$, where
	$t_{\mathrm{RD}}$ is the duration of ringdown (maximum amplitude to the end of the strain) and $T$ is the duration of the
	entire GW strain. 
%	\te{What is $t_{\mathrm{RD}}$? and how is it computed?}
} before using standard FFT routines to compute the fast Fourier transform.

@@

To evaluate how well the optimization procedure generalizes to waveforms that are
not in the training set, we evaluate the mismatch between the fine-tuned model
and an additional 526 NR waveforms in the SXS catalog i.e., the validation set.
We select waveforms that share the same part of the parameter space with the training
set, i.e., waveforms with negligible eccentricity (${e<2\times10^{-3}}$)
and precession (${\chi_{x,y}<5\times10^{-3}}$). Figure~\ref{fig:intrin_space}
shows how the training and validation waveforms are distributed in the
$q-\chi_{\rm PN}$ space.

@@

\begin{figure}[t]
	\script{q148.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/q148.pdf}
	\caption{Distributions of mismatches before and after optimization using the 
	$\mathcal{L}_{\mathrm{ave}}$ loss function. Mismatches are calculated using the training waveforms listed in 
	Tab.~\ref{tab:q148} and	are weighted with a constant PSD. 
	The dashed lines represent the median of the distributions, which decreased by 45.3\% during optimization.}
	\label{fig:q148}
\end{figure}
\begin{figure}[t]
	\script{q148_q1248_compare.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/q148_q1248_compare.pdf}
	\caption{Distributions of $\log_{10}$ difference in mismatch. The red
	distribution uses the training waveforms listed in
	Tab.~\ref{tab:q148} while the green distribution uses waveforms
	listed in Tab.~\ref{tab:q148}~and~\ref{tab:q1248}. Mismatches are calculated
	using a constant PSD with the loss function $\mathcal{L}_{\mathrm{ave}}$. 
	The dashed lines represent the median of the distributions, which decreased by an additional 10.8\% during optimization with the additional data.}
	\label{fig:q148_q1248_compare}
\end{figure}

Note that the performance of the IMRPhenomD model was initially tested using the
{\zdethp} weighted mismatch.\footnote{
	Note, however, that the mismatch was never directly used during the calibration process~\citep{khan2016frequency, khan2019phenomenological}.
} We would therefore like to examine whether using the {\zdethp} PSD in our loss function
could lead to an improved mismatch. Performing the optimization again, we find no significant difference
between the results using the two PSDs in the distribution of mismatches.

@@

We divided the parameter space into four regions to analyze the effect of the
recalibration procedure on each region separately
(Fig.~\ref{fig:all_quadrants}~and~\ref{fig:ps_q148_quadrant}). The training waveforms used for fitting in
this scenario are listed in Tab.~\ref{tab:quadrants} and the loss functions 
are calculated using a simple average of the mismatches, $\mathcal{L}_{\mathrm{ave}}$. The top-left and
bottom-right regions have limited data for $q>4$, hence the result is only valid up to $q\leq4$ and we take waveforms with $q\leq4$ as validation waveforms in these two regions. From top panel of Fig.~\ref{fig:ps_q148_quadrant}, we observe that
equal-spin waveforms lying on the diagonal improve the most. 
\kl{Both top-right and bottom-left region have good performance comparing with the original model. Especially for the bottom-left region, where the improvement is significantly better than optimizing all regions simultaneously. This suggests the ansatz are well-fitted to this part of the parameter space.
Waveforms above the diagonal also give good improvements when comparing with the original model and it has a similar improvement as optimizing all waveforms at once. Notice optimizing all regions gives a slightly better improvement. This shows that the ansatz can achieve a better fit using training waveforms with similar spins, even for waveforms with non-equal spin parameters.}
%The improvement above the diagonal is also significantly more than when optimizing using
%all waveforms at once. This can be seen clearly in the top panel of Fig.~\ref{fig:all_quadrants}, where we see
%that the optimized histogram shifts to lower mismatches.
\footnote{
	Note that the bottom-left panel of Fig.~\ref{fig:all_quadrants} specifically shows the
	distribution of mismatches for the top-left quadrant of Fig.~\ref{fig:ps_q148_quadrant}.
} 
On the other hand, the improvement below the diagonal is negligible, indicating that our method struggles to
further improve over the original IMRPhenomD waveform. 
Interestingly, we can see in the bottom-right region that the waveform does not show a significant change
to the overall mismatch (see bottom-right panel of Fig.~\ref{fig:all_quadrants}). 
This suggests that the IMRPhenomD waveform ansatz fits waveforms with
$\chi_1>0$ and $\chi_2<0$ less well. 
\kl{Overall, the improvement for both optimization schemes are similar (bottom panel of Fig.~\ref{fig:ps_q148_quadrant}). Although split region optimization uses more training waveforms, some of the waveforms, e.g. opposite spins, hinders the optimization procedure and hence gives a slightly worse result when comparing with all region optimization.}
